---
title: "Lab 1.Exercise"
authors:
- Adapted by EVL, FRC and ASP
date: "`r Sys.Date()`"
format:
    html: 
      toc: true
      toc-depth: 3
      code-fold: false
      fig-width: 8
      fig-height: 6
    pdf: default
knit:
  quarto:
    chunk_options:
      echo: true
      cache: false
      prompt: false
      tidy: true
      comment: NA
      message: false
      warning: false
    knit_options:
      width: 75
reference-location: margin
execute:
    echo: true
    message: false
    warning: false
bibliography: "../../StatisticalLearning.bib"
---

# Exercise

This problem involves the $OJ$ data set which is part of the ISLR2 package.

(a) Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.
(b) Use the `tree` package to fit a tree to the training data, with Purchase as the response and the other variables as predictors. Use the summary () function to produce summary statistics about the tree, and describe the results obtained. What is the training error rate? How many terminal nodes does the tree have?
(c) Type in the name of the tree object in order to get a detailed text output. Pick one of the terminal nodes, and interpret the information displayed.
(d) Create a plot of the tree, and interpret the results.
(e) Use the tree package to predict the response on the test data, and produce a confusion matrix comparing the test labels to the predicted test labels. What is the test error rate?
(f) Apply the cv.tree() function to the training set in order to determine the optimal tree size.
(g) Produce a plot with tree size on the $x$-axis and cross-validated classification error rate on the $y$-axis.
(h) Which tree size corresponds to the lowest cross-validated classification error rate?
(i) Produce a pruned tree corresponding to the optimal tree size obtained using cross-validation. If cross-validation does not lead to selection of a pruned tree, then create a pruned tree with five terminal nodes.
(j) Compare the training error rates between the pruned and unpruned trees. Which is higher?

Once you have solved the exercise, try to repeat it using another R package, either `rpàrt` or `caret`. Compare the results obtained and comment about the differences observed.

```{r}
library(dplyr)
```


# Solution using `tree`

```{r}
# Load the ISLR2 package and the OJ data set
library(ISLR2)
data(OJ)

# Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations
set.seed(1)
train_idx <- sample(nrow(OJ), 800)
OJ_train <- OJ[train_idx, ]
OJ_test <- OJ[-train_idx, ]

# Fit a tree to the training data, with Purchase as the response and the other variables as predictors
library(tree)
tree <- tree(Purchase ~ ., data = OJ_train)

# Use the summary() function to produce summary statistics about the tree, and describe the results obtained
summary(tree)
cat("Training error rate for unpruned tree:", tree$dev/sum(tree$frame$wt), "\n")
cat("Number of terminal nodes in unpruned tree:", tree$frame$var[tree$frame$var == "<leaf>"] %>% length, "\n")

# Create a plot of the tree
plot(tree)
text(tree, pretty = 0, cex=0.8)

# Predict the response on the test data, and produce a confusion matrix comparing the test labels to the predicted test labels
test_preds <- predict(tree, OJ_test, type = "class")
table(test_preds, OJ_test$Purchase)

# Calculate the test error rate
test_error_unpruned <- mean(test_preds != OJ_test$Purchase)
cat("Test error rate for unpruned tree:", test_error_unpruned, "\n")

# Apply the cv.tree() function of the tree package to the training set in order to determine the optimal tree size
cv <- cv.tree(tree, FUN = prune.misclass)

# Give the code to Produce a plot with tree size on the x-axis and cross-validated classification error rate on the y-axis
plot(cv$size, cv$dev, type = "b", xlab = "Tree Size", ylab = "Cross-Validated Error Rate")

# Use the tree package to Produce a pruned tree corresponding to the optimal tree size obtained using cross-validation
pruned_tree <- prune.misclass(tree, best = cv$size[which.min(cv$dev)])

# If cross-validation does not select a pruned tree, create a tree with five terminal nodes
if (is.null(pruned_tree)) {
  pruned_tree <- prune.misclass(tree, best = 8)
}

# Compare the training error rates between the pruned and unpruned trees
train_preds <- predict(tree, OJ_train, type = "class")
train_error_unpruned <- mean(train_preds != OJ_train$Purchase)
cat("Training error rate for unpruned tree:", train_error_unpruned, "\n")

train_preds <- predict(pruned_tree, OJ_train, type = "class")
train_error_pruned <- mean(train_preds != OJ_train$Purchase)
cat("Training error rate for pruned tree:", train_error_pruned, "\n")


```


# Solution using `rpart`


```{r}
# load the necessary packages
library(rpart)
library(rpart.plot)

# load the OJ dataset
data(OJ)

# split the data into training and test sets
set.seed(123)
train <- sample(nrow(OJ), 800)
train_data <- OJ[train, ]
test_data <- OJ[-train, ]

# fit a decision tree to the training data
tree_model <- rpart(Purchase ~ ., data = train_data, method = "class")

# summarize the tree model
summary(tree_model)

# plot the tree model
rpart.plot(tree_model, type = 0)

# make predictions on the test data
test_preds <- predict(tree_model, test_data, type = "class")

# create a confusion matrix
conf_matrix <- table(test_data$Purchase, test_preds)

# calculate the test error rate
test_error_rate <- 1 - sum(diag(conf_matrix)) / sum(conf_matrix)

# use cross-validation to find the optimal tree size
cv_results <- rpart.control(cp = seq(0.0001, 0.01, by = 0.0001))
cv_model <- rpart(Purchase ~ ., data = train_data, method = "class", control = cv_results)
plotcp(cv_model)

# prune the tree to the optimal size
optimal_cp <- cv_model$cptable[which.min(cv_model$cptable[, "xerror"]), "CP"]
pruned_model <- prune(tree_model, cp = optimal_cp)

# compare training error rates between the pruned and unpruned trees
train_preds <- predict(tree_model, train_data, type = "class")
train_error_rate_unpruned <- 1 - sum(diag(table(train_data$Purchase, train_preds))) / nrow(train_data)

train_preds_pruned <- predict(pruned_model, train_data, type = "class")
train_error_rate_pruned <- 1 - sum(diag(table(train_data$Purchase, train_preds_pruned))) / nrow(train_data)

cat("Training error rate (unpruned):", train_error_rate_unpruned, "\n")
cat("Training error rate (pruned):", train_error_rate_pruned, "\n")

```


# Solution using `caret`

```{r eval=FALSE}
# load the necessary packages
library(caret)

# load the OJ dataset
data(OJ)

# split the data into training and test sets
set.seed(123)
train_index <- createDataPartition(OJ$Purchase, p = 0.8, list = FALSE)
train_data <- OJ[train_index, ]
test_data <- OJ[-train_index, ]

# fit a decision tree to the training data
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
model <- train(Purchase ~ ., data = train_data, method = "rpart", trControl = ctrl)

# summarize the tree model
print(model$finalModel)

# plot the tree model
plot(model$finalModel, main = "Decision Tree")

# make predictions on the test data
test_preds <- predict(model, newdata = test_data)

# create a confusion matrix
conf_matrix <- table(test_data$Purchase, test_preds)

# calculate the test error rate
test_error_rate <- 1 - sum(diag(conf_matrix)) / sum(conf_matrix)

# use cross-validation to find the optimal tree size
tune_grid <- expand.grid(cp = seq(0.0001, 0.01, by = 0.0001))
tune_results <- train(Purchase ~ ., data = train_data, method = "rpart", trControl = ctrl, tuneGrid = tune_grid)

# prune the tree to the optimal size
pruned_model <- prune(tune_results$finalModel, cp = tune_results$bestTune$cp)

# compare training error rates between the pruned and unpruned trees
train_preds_unpruned <- predict(model, newdata = train_data)
train_error_rate_unpruned <- 1 - sum(diag(table(train_data$Purchase, train_preds_unpruned))) / nrow(train_data)

train_preds_pruned <- predict(pruned_model, newdata = train_data)
train_error_rate_pruned <- 1 - sum(diag(table(train_data$Purchase, train_preds_pruned))) / nrow(train_data)

cat("Training error rate (unpruned):", train_error_rate_unpruned, "\n")
cat("Training error rate (pruned):", train_error_rate_pruned, "\n")

```

# References



